Learning Network Size while Training with ShrinkNets

### 2018-02-09 paper/ICML/evaluation.tex

The goal of our evaluation is to show two main properties of
\textit{ShrinkNets}: they are exploring the space of possible sizes in a "smart"
way, and that as a result we obtain the smaller or equal size than
\textit{Static Networks}. We will then evaluate the performance benefits implied
by these gains in size.

\subsection{Datasets and Setup}

@@

Previous attempt at solving this problem focus on simple datasets and simple
architectures (limited to Fully connected). We believe that self-resizing
network really matter for problems where the architecture you need to solve them
involve so many layers that it is impractical to explore the space of possible
sizes. We considered ImageNet \cite{ILSVRC15} but the time required to train models is so long
that it would have not been possible to train enough and get statistically
significant results. This is why we picked \texttt{CIFAR10} \cite{Krizhevsky2009}. It is an
image classification dataset containing $60000$ color images $(3 \times 32
\times 32)$, belonging to $10$ different classes.

To solve this task we use the \texttt{VGG16} model \cite{Srivastava2014}. It is
constituted of alternating convolutional layers and \textit{MaxPool} layers
interleaved by \textit{BatchNorm} \cite{DBLP:journals/corr/IoffeS15} and
\textit{ReLU} \cite{Nair2010}  layers. The two last layers are Fully connected
layers separated by just a \textit{ReLU} activation function.

To turn it into a ShrinkNet we introduce \textit{Switch Layers} after each
\textit{BatchNorm} and each Fully connected (except the last one).

ShrinkNets assume that the starting size of the network is an upper bound on the
optimal size. We though that picking two times the recommended size for each
layer (that was designed for ImageNet), is a generous upper bound. For the
classification layers we use $5000$ neurons as an upper bound where the ImageNet
version uses $4096$ \gl{This is on the top of my head, need to be double checked}.

We assume no prior knowledge on the optimal batch size, learning rate,
$\lambda$ or weight decay ($\lambda_2$). This is why, we randomly sample them
from a range of reasonable values (\gl{should we make them explicit ?}).
Training is done using our library, based on \texttt{PyTorch} \cite{paszke2017automatic}, that
support dynamically resized layers. We train the network using gradient descent
and \textit{Adam} optimizer \cite{DBLP:journals/corr/KingmaB14}. We start with
the learning rate sampled randomly and every $5$ epochs of non improvement in
validation accuracy we divide the learning rate by $10$. We stop training after
$400$ epochs or when the learning rate is under $10^{-7}$ whichever comes
first. \gl{Should I also give the details about the removal strategy we used
  and the $\gamma$ and threshold I used ? because this is clearly getting
  boring} For each of the models we trained, we pick the epoch with the best
validation accuracy and report the corresponding testing accuracy. Because of
the nature of our method, it can happen that for network that are aggressively
compressed, the best validation accuracy is obtained early in the training,
before the size has converged. To be sure that accuracy measured corresponds to
the final shape and not the starting shape, we only consider the second half of
the training when picking the best epoch. For each model, we also measure the
total size, in number of floating point parameters, excluding the
\textit{Switch Layers} because as we saw in \cref{neuron_killing}, we can
get rid of them when training is done.

We want to compare against classical (\textit{Static}) networks. The number of
parameters that control the size is large: 13 for the size of convolutional
layers and $2$ for the fully connected ones. Without Shrinknets to help and fuse
all these parameters in a single $\lambda$ it is infeasible to sample reasonably
well a search space of that size. This is why we have to rely on the very well
known heuristic that the original VGG architecture (and many CNNs) \gl{try to
  find the paper that introduces this heuristic}. For \textit{Static Networks}
we sample the size between $0.1$ and $2$ times the size optimized for ImageNet.
We report the same numbers as we did for \textit{ShrinkNets} and we compare the
two distributions of models we obtain on the first plot of \cref{figure_CIFAR10}.

@@

The second dataset we use for evaluation the \texttt{COVERTYPE} \cite{Blackard:1998:CNN:928509} dataset. This dataset
contains $581012$ description of geographical area (elevation, inclination,
etc...) and the goal is to predict the type of forest growing there. We picked
this dataset for several reasons. First it is simple, such that we can reach good
accuracy with only a few layers. This important because we want to show that
\textit{ShrinkNets} find sizes as good as \textit{Static Networks}, even if we
are sampling the entire space of possible network sizes. The second reason is that Scardapane et al~\cite{Scardapane2017}
also perform their evaluation on this dataset, which allow us to compare the
results obtained by the two methods. \srm{what two methods?}

The only difference in terms of setup are minimal. We use the same architecture
as \cite{Scardapane2017}, i.e. three layers Fully Connected neural network with  no
\textit{Dropout} \cite{Srivastava2014} nor \textit{BatchNorm}. \gl{Should we say here that we
  don't expect Dropout to work here ? I could write an entire paragraph about it
  if needed}. In this case for the \textit{Static Networks}, we sample
independently sizes of the three different layers to explore all possible
architectures.

\subsection{Results of the random search procedure}

On the plot, the Pareto front for the two training methods clearly indicates
that not only \textit{ShrinkNets} find models that are as accurate as
\textit{Static Networks}, but they are more efficient in terms of size. Even
when we sample the entire size space (in the \texttt{COVERTYPE} case),
\textit{ShrinkNets} are better or equally as good as \textit{Static Networks}.
it has two implications. First it reinforces, our observation in [ref] that
showed that the models do not suffer from the non-convexity introduced by the
multiplication by the \textit{Switch Layer}. Secondly it could let us conjecture
that the parametrization from a $\lambda$ (real value) to the size of the
size of the network (vector) close to optimal. Indeed, if it was
not the case then eventually a randomly sampled size would have beaten
\textit{ShrinkNets} models. \gl{Too strong ?}

@@

We showed in the previous section that \textit{ShrinkNets} were able to find
more or at least as efficient as \textit{Static Networks}. In this experiment we
want to determine benefits of the smaller size we get.

@@

Regarding size, the key observation we can make is that size improvenets are
significant for the \texttt{CIFAR10}. In the range of accuracies we are
interested in, improvements in size go from 4x to 40x. On the \texttt{COVERTYPE}
dataset even if the compression ratio is always above 1, it hardly exceeds 3x,
except for very high accuracy where \textit{ShrinkNets} obtained some
particularly good solutions.

We observe that unlike local sparsity compression based method, our method
translates directly improvement in size to higher bandwidth at inference time.
In situations where inference is already incredibly fast (\texttt{COVERTYPE}
with a batch size of $1$ and \texttt{CIFAR10} with a batch size of $1$ on GPU),
improvements are dominted by noise. However, on machine learning frameworks
more focused on latency it should be possible to measure them.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{CIFAR10_VGG_summary}
\vspace*{-5mm}
\caption{\label{figure_CIFAR10} Summary of the result of random
search over the hyper-parameters the \texttt{CIFAR10} dataset}
\end{center}
\vspace*{-4mm}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{COVER_FC_summary}
\vspace*{-5mm}
\caption{\label{figure_COVER} Summary of the result of random
search over the hyper-parameters the \texttt{COVERTYPE} dataset
\srm{label the x axes on the bottom plots}
}
\end{center}
\vspace*{-4mm}
\end{figure}

@@

We sucessfully demonstrated that the architecture \textit{Shrinknets} converge
to make sense both in term of size and performance. And for a given accuracy
the size needed is significantly smaller than when we use the classic heursitc
we commonly use to size convolutional neural networks. It might be an
indication that this strategy is suboptimal and might need to be adjusted.
During our experimentations on simpler datasets like \texttt{MNIST}
\cite{Lecun1998} and \texttt{FashionMNIST} \cite{Xiao2017} we observed that
even if they have the same number of classes, input features, and output
distribution. For a fixed $\lambda$ \textit{ShrinkNets} converged to
considerably bigger networks in the case of \texttt{FashionMNIST}. This
evidence shows that final architecture not only depends on the output
distribution or shape of the data but actually reflects the dataset, indeed we
know that \texttt{MNIST} is a much easier problem than \texttt{FashionMNIST}.

We provide two examples of architectures learned by \textit{ShrinkNet}. One
for the best test available test accuracy and a slightly underperforming but
significantly smaller that it's equivalent \textit{Static Network}. For
these two architectures we show the size at different time during training to
give a sense how the convergence happen.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{size_evolution}
\vspace*{-5mm} \caption{\label{network_size_evolution} Evolution of the size of
  each layer over time (lighter: beginning, darker: end). On top a very large
  network performing $92.07\%$, at the bottom a simpler model with $90.5\%$
  accuracy. 
} \end{center}
\vspace*{-4mm}
\end{figure}

>>>>

The goal of our evaluation is to explore:
\begin{itemize}
	\item
Whether, by varying $\lambda$, \textit{ShrinkNets} can efficiently explore (in terms of number of training runs)  the spectrum of high-accuracy models from small to large, on both CNNs and fully connected networks.  Our results show that, for each network size, we obtain models that perform as well or better than 
\textit{Static Networks}, trained via traditional hyperparameter optimization.

\item Whether, because these  smaller networks are dense, they result in improved inference times on both CPUs and GPUs.

\item Whether the ShrinkNets approach results in network architectures that are substantially different than the best network architecutres (in terms of relative number of neurons per layered) identified in the literature.

\item \srm{something else?}
\end{itemize}

\subsection{Performance vs. Traditional Methods}

We evaluate ShrinkNets on two datasets:  \texttt{CIFAR10} and \texttt{COVERTYPE}.  In this section, we report how we employed ShrinkNets on these datasets and compare our results to 
training without ShrinkNets.

@@

Previous research~\cite{Scardapane2017} that attempts to optimize network size during training
 focused on simple datasets and simple (i.e., fully connected)
architectures. We believe that self-resizing
network are most important for problems where the architecture 
involves many layers, such that that it is impractical to explore the space of all possible
layer-size configurations. For that reason, we picked \texttt{CIFAR10} \cite{Krizhevsky2009}. It is an
image classification dataset containing $60000$ color images $(3 \times 32
\times 32)$, belonging to $10$ different classes, running on  the \texttt{VGG16} network \cite{Srivastava2014}. VGG16
consists of alternating convolutional layers and \textit{MaxPool} layers
interleaved by \textit{BatchNorm} \cite{DBLP:journals/corr/IoffeS15} and
\textit{ReLU} \cite{Nair2010}  layers. The two last layers are fully connected
layers separated by a \textit{ReLU} activation function.

\srm{keep this?}
 Although ShrinkNets supports larger networks and datasets as well (i.e., ImageNet \cite{ILSVRC15}),  the time required to such train models and get statistically significant result was prohibitive -- especially when comparing against classical hyperparameter search over network architecture.

We applied ShrinkNets to the VGG16 network by adding \textit{Switch Layers} after each
\textit{BatchNorm} layer and each fully connected layer (except the last).
Recall that ShrinkNets assume that the starting size of the network is an upper bound on the
optimal size. For this reason, we simply started with a network with double the recommended size for each
layer as an upper bound (this is larger than what ImageNet, for example uses). Thus, for the
classification layers we use $5000$ neurons as a starting limit (ImageNet
 uses $4096$ \gl{This is on the top of my head, need to be double checked}).

We assume no prior knowledge on the optimal batch size, learning rate,
$\lambda$ or weight decay ($\lambda_2$). Instead, we trained a number of models,
 randomly and independently selecting the values of these parameters 
from a range of reasonable values (\gl{should we make them explicit ?} \srm{yes}).
We trained using our modified \texttt{PyTorch} \cite{paszke2017automatic} library that
supports dynamically resized layers (our code is available at \srm{add anonymized link}). Training is done using gradient descent and the
 \textit{Adam} optimizer \cite{DBLP:journals/corr/KingmaB14}. Specifically, 
 we start with
the learning rate sampled randomly; for every $5$ epochs of non-improvement in
validation accuracy we divide the learning rate by $10$. We stop training after
$400$ epochs or when the learning rate is under $10^{-7}$, whichever comes
first.  \gl{Should I also give the details about the removal strategy we used
  and the $\gamma$ and threshold I used ? because this is clearly getting
  boring} \srm{probably not necessary} 
  For each of the models we trained, we pick the epoch with the best
validation accuracy and report the corresponding testing accuracy. Because of
the nature of our method, it can happen that for networks that are aggressively
compressed, the best validation accuracy is obtained early in  training,
before the size has converged. To be sure that accuracy measured corresponds to
the final shape and not the starting shape, we only consider the second half of
the training when picking the best epoch. For each model, we also measure the
total size, in terms of number of floating point parameters, excluding the
\textit{Switch Layers} because as described in \cref{neuron_killing}, these are eliminated
after training.

We compare against classical (\textit{Static}) networks. In such networks, the number of
parameters that control the size is large: 13 parameters for the convolutional
layers and $2$ for the fully connected layers. ShrinkNets effectively fuse
all these parameters in a single $\lambda$, but in conventional architectures
where all of these parameters are free, it is infeasible to obtain a reasonable
sample of a search space of this size. For this reason, we rely on the conventional
 heuristic that the original VGG architecture (and many CNNs) \gl{try to
  find the paper that introduces this heuristic} user, where \srm{describe what the heuristic is -- doubling of neurons up to apoint?} For \textit{Static Networks}
we sample the size between $0.1$ and $2$ times the size \srm{size of what?} optimized for ImageNet. \srm{Why are we using ImageNet and not CIFAR10 as the comparison point?}
We report the same numbers as we did for \textit{ShrinkNets} and we compare the
two distributions.  
\srm{I worry that this method of selecting network sizes for Static will of course constrain it to larger sizes and worse perforamance.  What if we just randomly sample networks?  Would it look terrible?}

The results are shown in the top figure of \cref{figure_CIFAR10}. Here, the blue dots show the models produced by ShrinkNets, and the orange dots show the models produced for the static networks.  For each model, we plot its accuracy and model size. The lines show the Pareto frontier of models in each of the two optimization settings. From the figure, it's apparent that ShrinkNets much more effectively explores the trade-off between model size and accuracy.
 \srm{Would be good to show Lambda values on the figure somehow.  }
 \srm{I thought we were going to summarize with distributional plots instead of Pareto frontiers?}
 \srm{Summarize more quantitatively -- best model performance, model performance drop off for a factor of 10 reduction in seize, etc. so that we can repeat a quantitative claim in the intro, i.e.,
 ``Note that the best performing ShrinkNets models has 9x.x\% accuracy while the best static model has 9x.x\% accuracy, while the ShrinkNets model is x times smaller.  In addition, if we give up just 1\% error, ShrinkNets finds a model that is z times smaller.'' (If we do this, would be good to highlight these points in the figure somehow)}

@@

The second dataset we use for evaluation is the \texttt{COVERTYPE} \cite{Blackard:1998:CNN:928509} dataset. This dataset
contains $581012$ descriptions of geographical area (elevation, inclination,
etc...) and the goal is to predict the type of forest growing in each area. We picked
this dataset for several reasons. First it is simple, such that we can reach good
accuracy with only a few fully-connected layers. This important because we want to show that
\textit{ShrinkNets} find sizes as good as \textit{Static Networks}, even if we
are sampling the entire space of possible network sizes. The second reason is that Scardapane et al~\cite{Scardapane2017}
also perform their evaluation on this dataset, which allow us to compare the
results obtained by our method with the method in ~\cite{Scardapane2017}.

The experimental setup on this dataset is similar to {\tt CIFAR10}. We use the same architecture
as \cite{Scardapane2017}, i.e., three fully-connected layers network with  no
\textit{Dropout} \cite{Srivastava2014} or \textit{BatchNorm}. \gl{Should we say here that we
  don't expect Dropout to work here ? I could write an entire paragraph about it
  if needed}. In this case, for the \textit{Static Networks}, we 
independently samples the sizes of the three different layers to explore all possible
architectures.

The results are shown in the top figure of \cref{figure_COVER}, with the two optimization methods plotted as before. Here, {\it Static} method finds models that perform well at a variety of sizes, because it is able to explore the entire parameter space.  This is as expected;  the fact that ShrinkNets performs as well as the Static indicates that ShrinkNets is doing an effective job of exploring the parameter space using just the single $\lambda$ parameter.  
 \srm{Summarize more quantitatively -- best model performance, model performance drop off for a factor of 10 reduction in seize, etc. so that we can repeat a quantitative claim in the intro, i.e.,
 ``Here, the best performing ShrinkNets models has 9x.x\% accuracy while the best static model has 9x.x\% accuracy, while the ShrinkNets model is x times smaller.  Further, note that ShrinkNets finds a model with 9x.x\% accuracy that is y times than the best performing model. 
 Also, as on CIFAR10, if we give up just 1\% error, ShrinkNets finds a model that is z times smaller.'' (If we do this, would be good to highlight these points in the figure somehow)}

\subsubsection{Discussion}

\srm{we may not need to keep this unless we can think of something more to say}

The Pareto frontier in the two figures shows that 
that  \textit{ShrinkNets} is able find models that both as accurate and smaller than
\textit{Static Networks}.
when we sample the entire size space (in the \texttt{COVERTYPE} case),
\textit{ShrinkNets} are better or equally as good as \textit{Static Networks}.
This has two implications. First it reinforces our observation \srm{our observation?} in [ref] that
showed that the models do not suffer from the non-convexity introduced by the
multiplication by the \textit{Switch Layer}. Secondly it could let us conjecture
that the parametrization from a $\lambda$ (real value) to the size of the
size of the network (vector) close to optimal. Indeed, if it was
not the case then eventually a randomly sampled size would have beaten
\textit{ShrinkNets} models. \gl{Too strong ?} \srm{I think this is too strong -- we have no evidence this is true, do we?}

@@

We showed in the previous experiments that the \textit{ShrinkNets} approach is able to find
networks with accuracy equal to or better than \textit{Static Networks}. In this experiment we
want to determine benefits of the smaller size we get.

@@

Regarding size, the key observation we can make is that size improvements are
significant for the \texttt{CIFAR10}. In the range of accuracies we are
interested in, improvements in size go from 4x to 40x. On the \texttt{COVERTYPE}
dataset even if the compression ratio is always above 1, it hardly exceeds 3x,
except for very high accuracy where \textit{ShrinkNets} obtained some
particularly good solutions.

We observe that unlike local sparsity compression based method, our method
translates directly improvement in size to higher bandwidth at inference time.
In situations where inference is already incredibly fast (\texttt{COVERTYPE}
with a batch size of $1$ and \texttt{CIFAR10} with a batch size of $1$ on GPU),
improvements are dominated by noise. However, on machine learning frameworks
more focused on latency it should be possible to measure them.

@@

\begin{figure*}[t]\centering
\begin{minipage}{2.7in}
\begin{center}
\includegraphics[width=\columnwidth]{CIFAR10_VGG_summary}
\vspace*{-5mm}
\caption{\label{figure_CIFAR10} Summary of the result of random
search over the hyper-parameters the \texttt{CIFAR10} dataset}
\end{center}
\vspace*{-4mm}
\end{minipage}
\begin{minipage}{2.7in}
\begin{center}
\includegraphics[width=\columnwidth]{COVER_FC_summary}
\vspace*{-5mm}
\caption{\label{figure_COVER} Summary of the result of random
search over the hyper-parameters the \texttt{COVERTYPE} dataset
\srm{label the x axes on the bottom plots}
}
\end{center}
\vspace*{-4mm}
\end{minipage}
\end{figure*}

@@

The previous experiments showed that \textit{ShrinkNets} finds architectures
the effectively explore the frontier of model size and accuracy. Further, for
a  given accuracy,
the size needed is significantly smaller than when we use the \srm{name it} heuristic
 commonly used to size convolutional neural networks. This suggests that this
 conventional heuristic may not in fact be optimal, especially when looking for 
 smaller models. 
Empirically we observed this to often be the case.  For example, 
during our experimentations on the \texttt{MNIST}
\cite{Lecun1998} and \texttt{FashionMNIST} \cite{Xiao2017} datasets (not reported here due to space constraints), we observed that
even though these datasets have the same number of classes, input features, and output
distributions,  for a fixed $\lambda$ \textit{ShrinkNets} converged to
considerably bigger networks in the case of \texttt{FashionMNIST}. This
evidence shows that optimal architecture not only depends on the output
distribution or shape of the data but actually reflects the dataset.  This makes sense,
as  \texttt{MNIST} is a much easier problem than \texttt{FashionMNIST}.

To illustrate this point on a larger dataset, 
we show two examples of architectures learned by \textit{ShrinkNets} in Figure~\ref{fig:network_size_evolution}.  The top plot shows the model with the best
 test accuracy, with \srm{identical performance to the best static network?}; 
 the bottom shows a network that slightly under-performs the best in terms of accuracy but is
significantly smaller that the best equivalent \textit{Static Network} \srm{how much smaller?}.
In the plot, the dashed line shows the number of neurons in each layer of the original VGG net, and the shaded regions show the size of the ShrinkNet as it converges (with the darkest region representing the fully converged network).  Observe that the final network that is trained looks quite different in the two cases, with the optimal performing network appearing similar to the original VGG net, whereas the shrunken network allocates many fewer neurons to the middle layers, and then additional neurons to the final fewer layers.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.7\columnwidth]{size_evolution}
\vspace*{-5mm} 
\caption{ Evolution of the size of
  each layer over time (lighter: beginning, darker: end). On top a very large
  network performing $92.07\%$, at the bottom a simpler model with $90.5\%$
  accuracy. 
} 
\label{fig:network_size_evolution}
\end{center}
\vspace*{-4mm}
\end{figure}
